Fixed-Parameter Tractable Sampling for RNA Design with Multiple Target Structures

### 2018-04-25 RNARedPrint-Bioinfo2018.tex
Mess with targeting figure and energy notation

\newcommand{\ourprog}{\Software{RNARedPrint}}

@@

\newcommand{\RNAblueprint}{{\tt \bfseries\color{blue}RNABlueprint}}

@@

One can interpret the valid sequences for $\R$ as colorings of
$G_{\R}$ in a slightly modified graph coloring variant, where the
colors (from $\Sigma$) assigned to adjacent vertices of $G_{\R}$ must
constitute valid base pairs.
%
We define the energy of a sequence $S$ based on the set of structures
$\R$ as $E_\R(S) \in \mathbb{R}\cup\{\infty\}$. In
our setting, the energy $E_\R(S)$ is additively composed of
the energies of the single RNA structures in an RNA energy model, as
well as sequence-dependent features like \GCb\%.  Furthermore note
that $E_\R(S)$ is finite iff $S$ is valid for each structure
$R_1,\dots,R_k$.

At the core of this work, we study the computation of partition
functions over sequences.\smallskip\\
\textbf{Central problem (Partition function over sequences).}
  Given an energy function $E$ and a set $\R$ of structures of length $n$, compute the
  partition function
  \begin{equation}
    \label{eq:mainproblem}
    \partfun{E_\R} = \sum_{S\in\S_n} \exp(-\beta E_\R(S)),
  \end{equation}
  where $\beta$ denotes the inverse pseudo-temperature, and $\S_n$ the set of sequences of length $n$.

%We solve the problem efficiently for any fixed parameter $w$, whose value reflects the intrinsic complexity of the structures and energy function $E_\R$.
%
  As we elaborate in subsequent sections, our approach relies on
  breaking down the energy function $E_\R(S)$ into additive
  components, each depending on only few sequence positions. Given
  $\R$, we express $E_\R(S)$ as the sum of energy contributions $f(S)$
  over a set $\F$ (of functions $f:\S_n\to\real$),
  s.t.~$E_\R(S)=\sum_{f\in\F} f(S)$. This captures realistic RNA
  energy models---including nearest neighbor models,
  e.g.~\citet{Turner2009}, and even pseudoknot models,
  e.g. \citet{Andronescu2010}, while bounding the dependencies to
  sequence positions introduced by each single $f$. Formally, define
  the \Def{dependencies} $\dep(f)$ of $f$ as the minimum set of sequence
  positions $\I\subseteq\{1,\dots,n\}$, where $f(S)=f(S')$ for all
  sequences $S$ and $S'$ that agree at the positions in $\I$.

@@

\subsection{Targeting Turner energies and \GCb\%}
We implemented our Boltzmann sampling strategy (Algorithms
\ref{alg:pf} and \ref{alg:sampling}), to sample valid sequences for given
target structures and weights $\pi_1,\dots,\pi_n$.
Moreover, we used multi-dimensional Boltzmann sampling, as described in Section~\ref{sec:multiBoltzmann}, to target specific energies and \GCb\%.  Our tool \ourprog{} evaluates energies according to the base
pair energy model or a stacking energy model $E_{\mathcal{S}}$, whose parameters 
were fitted to best approximate Turner energies (Supp. Mat.~\ref{appsec:modelparameters}).
%
%\paragraph{Implementation of multi-dimensional Boltzmann sampling}
%As suggested in Section~\ref{sec:multiBoltzmann}, we heuristically optimize weights to target specific energies and \GCb\% in an iterative procedure (based on the core sampling algorithm).
%Specifically, energy weights are initialized as $w_\ell=e^{\frac{1}{RT}}$; the gc weight as $w_{gc}=1$. For each iteration, $\psi\cdot n$ sequences are sampled with the current weights. Then, we determine the mean energies $E(S;R_\ell)$ and the mean GC content over the sampled sequences. For the next iteration, the weights are updated by the rules $w_\ell = w_\ell \gamma^{mean(E(S;R_\ell))-E_\ell}$ and $w_{gc} = w_{gc}\frac{GC_{target}}{mean(GC)}$. During this procedure, we collect   in a set $P$ all the sequences exhibiting energies within $[E_1-\delta,E_1+\delta]\times [E_2-\delta,E_2+\delta]\ldots$ and and GC-content within $[GC_{target}-\idelta',GC_{target}+\delta']$.
%If $|P| > n$ or an maximum amount of iterations is reached, the sequences in $P$ are returned.
%

To capture the realistic Turner model $E_{\mathcal{T}}$ while keeping the computational cost in check, we exploited the excellent linear fit  of $E_{\mathcal{T}}$  with the fitted stacking model $E_{\mathcal{S}}$ (supp. Fig~\ref{fig:training-cor}). Namely, we observed a structure-specific affine dependency between the Turner and stacking energy models, so that $E_{\mathcal{T}}(S;R) \approx \gamma\cdot E_{\mathcal{S}}(S;R) + \delta$ for any structure $R$ and sequence $S$. We learned the $(\gamma,\delta)$ parameters from a set of sequences generated with homogenous weights $w=e^{\beta}$, tuning only the \GCb\% to a predetermined value.  Finally, we adjusted the targeted energies within our stacking model to $E_{\mathcal{S}}^{\star} = (E_{\mathcal{T}}^{\star}- \delta)/\gamma$ in order to reach, on average, the right targeted energy $E_{\mathcal{T}}^{\star}$ in the Turner model.

Fig.~\ref{fig:energydist} illustrates our above strategy. Fig.~\ref{fig:energydist}A
shows the relative distance between $E_{\mathcal{T}}$  and $E_{\mathcal{S}}$, observed for instances of the \texttt{Modena} benchmark. For each structure, $n=1\,000$ sequences were sampled, while targeting a ${\sf GC}\%$ of $50\%$, and reported the difference between the expected and sampled energies for both the Turner model and its approximation using a stacking model. 
Fig.~\ref{fig:energydist}B shows the Turner energy, for a single instance of the 3-structures \texttt{Modena}  benchmark, using different weights benchmark. These results demonstrate that our multidimensionnal Boltzmann sampling strategy can accurately and, to a large extent, independently shift the Turner energies of sampled sequences to prescribed targets.

@@

\Final{
Extensions:
\begin{enumerate}
  \item What if the input graph is not bipartite? \\
  Extract a minimal subset of edges such that new graph is bipartite!\\
  $\to{}$ Graph bipartization (cf implementation by Hüffner, JGAA 2009)
  \item Is the optimization version in the \emph{weighted base-pair model} already {\sf NP}-hard?
  \item Does it help to know that there are cliques of size $k$ during the tree decomposition?\\
  If anything, it provides a lower bound on tree decomposition (helps branch and bound algorithms)
  \item What if some degree of freedom is allowed in the structure definition?\\ Can we still use tree decomposition?\\
  This looks like a very reasonable first step towards the definition of an {\em optimal} DP decomposition for RNA folding (with Pseudoknots) or RNA-RNA interactions\ldots
  \item Structures can also be penalized, but they must be specified explicitly.\\
  However, any competing structure must consist of helices, and it is possible to penalize each of the $\Theta(n^2)$ helices of any given length $L$. What would be the impact on treewidth?
  \item More general energy  models $\to$ Turner. (Pseudoknots)
  For this description, we assume that all input structures are non-crossing, while the framework would otherwise support crossing input structures as well as according energy models (e.g. the Hotknots model~\citep{Ren2005}) by considering special pseudoknot loops.
\end{enumerate}\label{lbl:theend}
}

>>>>

@@

\newcommand{\RNAblueprint}{{\tt \bfseries{} RNA\textcolor{blue!70!black}{Blue}print}}
\newcommand{\ourprog}{{\tt \bfseries{}RNA\textcolor{red!70!black}{Red}Print}}

\newcommand{\Energy}[2]{\textcolor{green!70!black}{E(#2;#1)}}

\newcommand{\EnergyTurner}{E_{\text{T}}}
\newcommand{\EnergyStacking}{E_{\text{st}}}

@@

One can interpret the valid sequences for $\R$ as colorings of
$G_{\R}$ in a slightly modified graph coloring variant, where the
colors (from $\Sigma$) assigned to adjacent vertices of $G_{\R}$ must
constitute valid base pairs.
%
The \Def{energy} of a sequence $S$ is defined by an \Def{energy
  function} $E$---defined for set of structures $\R$ and mapping
sequences to $\mathbb{R}\cup\{\infty\}$---as
$\Energy{\R}{S} \in \mathbb{R}\cup\{\infty\}$. In our setting, the
energy $\Energy{\R}{S}$ is typically chosen as a linear combination of
the energies of the single RNA structures (in some RNA energy model
that defines $E(S,R) \in \mathbb{R}\cup\{\infty\}$ for each
sequence $S$ and structure
$R$), as well as sequence-dependent features like \GCb\%. Formally, we
will require that
$\Energy{\R}{S}$ can be expressed as sum over functions values, where
each function depends on the nucleotides at certain positions of
$S$ (see below). Furthermore note that
$\Energy{\R}{S}$ is finite iff
$S$ is valid for each structure $R_1,\dots,R_k$.

At the core of this work, we study the computation of partition
functions over sequences.\smallskip\\
\textbf{Central problem (Partition function over sequences).}
  Given an energy function $E$ and a set $\R$ of structures of length $n$, compute the
  partition function
  \begin{equation}
    \label{eq:mainproblem}
    \partfun{E,\R} = \sum_{S\in\S_n} \exp(-\beta \Energy{\R}{S}),
  \end{equation}
  where $\beta$ denotes the inverse pseudo-temperature, and $\S_n$ the set of sequences of length $n$.

%We solve the problem efficiently for any fixed parameter $w$, whose value reflects the intrinsic complexity of the structures and energy function $E_\R$.
%
  As we elaborate in subsequent sections, our approach relies on
  breaking down the energy function $\Energy{\R}{S}$ into additive
  components, each depending on only few sequence positions. Given
  $\R$, we express $\Energy{\R}{S}$ as the sum of energy contributions $f(S)$
  over a set $\F$ (of functions $f:\S_n\to\real$),
  s.t.~$\Energy{\R}{S}=\sum_{f\in\F} f(S)$. This captures realistic RNA
  energy models---including nearest neighbor models,
  e.g.~\citet{Turner2009}, and even pseudoknot models,
  e.g. \citet{Andronescu2010}, while bounding the dependencies to
  sequence positions introduced by each single $f$. Formally, define
  the \Def{dependencies} $\dep(f)$ of $f$ as the minimum set of sequence
  positions $\I\subseteq\{1,\dots,n\}$, where $f(S)=f(S')$ for all
  sequences $S$ and $S'$ that agree at the positions in $\I$.

@@

\subsection{Targeting Turner energies and \GCb\%}
We implemented our Boltzmann sampling strategy (Algorithms
\ref{alg:pf} and \ref{alg:sampling}), to sample valid sequences for given
target structures and weights $\pi_1,\dots,\pi_n$.
Moreover, we used multi-dimensional Boltzmann sampling, as described in Section~\ref{sec:multiBoltzmann}, to target specific energies and \GCb\%.  Our tool \ourprog{} evaluates energies according to the base
pair energy model or a stacking energy model $\EnergyStacking$, whose parameters 
were fitted to best approximate Turner energies (Supp. Mat.~\ref{appsec:modelparameters}).
%
%\paragraph{Implementation of multi-dimensional Boltzmann sampling}
%As suggested in Section~\ref{sec:multiBoltzmann}, we heuristically optimize weights to target specific energies and \GCb\% in an iterative procedure (based on the core sampling algorithm).
%Specifically, energy weights are initialized as $w_\ell=e^{\frac{1}{RT}}$; the gc weight as $w_{gc}=1$. For each iteration, $\psi\cdot n$ sequences are sampled with the current weights. Then, we determine the mean energies $E(S;R_\ell)$ and the mean GC content over the sampled sequences. For the next iteration, the weights are updated by the rules $w_\ell = w_\ell \gamma^{mean(E(S;R_\ell))-E_\ell}$ and $w_{gc} = w_{gc}\frac{GC_{target}}{mean(GC)}$. During this procedure, we collect   in a set $P$ all the sequences exhibiting energies within $[E_1-\delta,E_1+\delta]\times [E_2-\delta,E_2+\delta]\ldots$ and and GC-content within $[GC_{target}-\idelta',GC_{target}+\delta']$.
%If $|P| > n$ or an maximum amount of iterations is reached, the sequences in $P$ are returned.
%

To capture the realistic Turner model $\EnergyTurner$ while keeping the computational cost in check, we exploited the excellent linear fit  of $\EnergyTurner$  with the fitted stacking model $\EnergyStacking$ (supp. Fig~\ref{fig:training-cor}). Namely, we observed a structure-specific affine dependency between the Turner and stacking energy models, so that $\EnergyTurner(S;R) \approx \gamma\cdot \EnergyStacking(S;R) + \delta$ for any structure $R$ and sequence $S$. We learned the $(\gamma,\delta)$ parameters from a set of sequences generated with homogenous weights $w=e^{\beta}$, tuning only the \GCb\% to a predetermined value.  Finally, we adjusted the targeted energies within our stacking model to $\EnergyStacking^{\star} = (\EnergyTurner^{\star}- \delta)/\gamma$ in order to reach, on average, the right targeted energy $\EnergyTurner^{\star}$ in the Turner model.

Fig.~\ref{fig:energydist} illustrates our above strategy. Fig.~\ref{fig:energydist}A
shows the relative distance between $\EnergyTurner$  and $\EnergyStacking$, observed for instances of the \texttt{Modena} benchmark. For each structure, $n=1\,000$ sequences were sampled, while targeting a ${\sf GC}\%$ of $50\%$, and reported the difference between the expected and sampled energies for both the Turner model and its approximation using a stacking model. 
Fig.~\ref{fig:energydist}B shows the Turner energy, for a single instance of the 3-structures \texttt{Modena}  benchmark, using different weights benchmark. These results demonstrate that our multidimensionnal Boltzmann sampling strategy can accurately and, to a large extent, independently shift the Turner energies of sampled sequences to prescribed targets.

@@

\Final{
Extensions:
\begin{enumerate}
  \item What if the input graph is not bipartite? \\
  Extract a minimal subset of edges such that new graph is bipartite!\\
  $\to{}$ Graph bipartization (cf implementation by Hüffner, JGAA 2009)
  \item Is the optimization version in the \emph{weighted base-pair model} already {\sf NP}-hard?
  \item Does it help to know that there are cliques of size $k$ during the tree decomposition?\\
  If anything, it provides a lower bound on tree decomposition (helps branch and bound algorithms)
  \item What if some degree of freedom is allowed in the structure definition?\\ Can we still use tree decomposition?\\
  This looks like a very reasonable first step towards the definition of an {\em optimal} DP decomposition for RNA folding (with Pseudoknots) or RNA-RNA interactions\ldots
  \item Structures can also be penalized, but they must be specified explicitly.\\
  However, any competing structure must consist of helices, and it is possible to penalize each of the $\Theta(n^2)$ helices of any given length $L$. What would be the impact on treewidth?
  \item More general energy  models $\to$ Turner. (Pseudoknots)
  For this description, we assume that all input structures are non-crossing, while the framework would otherwise support crossing input structures as well as according energy models, e.g.{} the Hotknots model of~\citet{Ren2005}, by considering special pseudoknot loops.
\end{enumerate}\label{lbl:theend}
}
